Preventing Runtime Errors of Redis at Compile Time

### 2016-06-11 aplas16/sections/Introduction.tex
Created a directory for aplas16.

\section{Indexed Monads}
\label{sec:indexed-monads}

Redis commands are \emph{actions}.
 We could capture the effects caused by an action, by expressing the states it
 affects, before and after. That is, the \emph{preconditions} and
 \emph{postconditions} of an action. In such way, we could also impose
 constraints on the preconditions.

\emph{Indexed monads} (or \emph{monadish},
 \emph{parameterised monad})\cite{indexedmonad}
 can be used\cite{typefun,staticresources} to model such preconditions and
 postconditions in types. An indexed monad is a type constructor that takes
 three arguments: an initial state, a final state, and the type of a value that
 an action computes, which can be read like a Hoare triple\cite{kleisli}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Class \ensuremath{\Conid{IMonad}} comes with two operations: \ensuremath{\Varid{unit}} for identities and \ensuremath{\Varid{bind}} for
compositions, as in monads.

We define a new datatype \ensuremath{\Conid{Popcorn}}, which is basically just
 the context \ensuremath{\Conid{Redis}} indexed with more information in types.
 We make it an instance of \ensuremath{\Conid{IMonad}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The first and second argument of type \ensuremath{\Conid{Popcorn}} is where the
dictionaries going to be.

To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to
 \ensuremath{\Varid{unPopcorn}} to get an ordinary Hedis program back,
 with type information erased.

\subsection{\text{PING}: the first attempt}

In Redis, \text{PING} does nothing but replies with
 \text{PONG} if the connection is alive. In Hedis,
 \ensuremath{\Varid{ping}} has type:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Now with \ensuremath{\Conid{Popcorn}}, we could make our own version of
\ensuremath{\Varid{ping}}\footnotemark

\footnotetext{\ensuremath{\Varid{ping}} from Hedis is qualified with
\ensuremath{\Conid{Hedis}} to prevent function name clashing in our code.}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Popcorn}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The dictionary \ensuremath{\Varid{xs}} in the type remains unaffected after the
 action, because \ensuremath{\Varid{ping}} does not affect any key-type
 bindings. To encode other commands that modifies key-type bindings, we need
 type-level functions to annotate those effects on the dictionary.

>>>>

\section{Introduction}
\label{sec:introduction}

\subsection{Redis}

Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
database, cache and message broker. A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with. Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.

For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
\text{some-set} abd \text{another-set} are
both associated to a set. The two call to command \text{SADD} respectively adds three and two values to the two sets, before \text{SINTER} takes their intersection:
\begin{tabbing}\tt
~redis\char62{}~SADD~some\char45{}set~a~b~c\\
\tt ~\char40{}integer\char41{}~3\\
\tt ~redis\char62{}~SADD~another\char45{}set~a~b\\
\tt ~\char40{}integer\char41{}~2\\
\tt ~redis\char62{}~SINTER~some\char45{}set~another\char45{}set\\
\tt ~1\char41{}~\char34{}a\char34{}\\
\tt ~2\char41{}~\char34{}b\char34{}
\end{tabbing}

Note that the keys \text{some-set} and \text{another-set}, if not existing before the call to \text{SADD}, are created on site. The call to
\text{SADD} returns the size of the set after
completion of the command.

\subsection{Hedis}

Many third party library exist to allow general purpose programmings
to access Redis databases through its TCP protocol. The most popular
such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.

The following program is how the previous example looks like in
Hedis:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 another-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34},\text{\tt \char34 c\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The function \ensuremath{\Varid{sadd}} takes a key and a list of values as
 arguments, and returns an \ensuremath{\Conid{Integer}} on success, or
 returns a \ensuremath{\Conid{Reply}}, a low-level representation of replies
 from the Redis server, in case of failures. All wrapped in
 \ensuremath{\Conid{Redis}}\footnotemark, the context of command execution.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{sadd}{}\<[7]%
\>[7]{}\mathbin{::}\Conid{ByteString}{}\<[27]%
\>[27]{}\mbox{\onelinecomment  key}{}\<[E]%
\\
\>[7]{}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]{}\<[27]%
\>[27]{}\mbox{\onelinecomment  values}{}\<[E]%
\\
\>[7]{}\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Note that keys and values, being nothing but binary strings in Redis,
are represented using Haskell \ensuremath{\Conid{ByteString}}.
Values of other types must be encoded as \ensuremath{\Conid{ByteString}}s before being written to the database,
and decoded after being read back.

\footnotetext{Hedis provides another kind of context, \ensuremath{\Conid{RedisTx}}, for \emph{transactions}, united with \ensuremath{\Conid{Redis}} under the class of \ensuremath{\Conid{RedisCtx}}. For brevity, we demonstrate only \ensuremath{\Conid{Redis}} in this paper. }

### 2016-06-11 aplas16/sections/Motivation.tex
Created a directory for aplas16.

\section{Introduction}
\label{sec:introduction}

\subsection{Redis}

Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
database, cache and message broker. A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with. Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.

For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
\text{some-set} abd \text{another-set} are
both associated to a set. The two call to command \text{SADD} respectively adds three and two values to the two sets, before \text{SINTER} takes their intersection:
\begin{tabbing}\tt
~redis\char62{}~SADD~some\char45{}set~a~b~c\\
\tt ~\char40{}integer\char41{}~3\\
\tt ~redis\char62{}~SADD~another\char45{}set~a~b\\
\tt ~\char40{}integer\char41{}~2\\
\tt ~redis\char62{}~SINTER~some\char45{}set~another\char45{}set\\
\tt ~1\char41{}~\char34{}a\char34{}\\
\tt ~2\char41{}~\char34{}b\char34{}
\end{tabbing}

Note that the keys \text{some-set} and \text{another-set}, if not existing before the call to \text{SADD}, are created on site. The call to
\text{SADD} returns the size of the set after
completion of the command.

\subsection{Hedis}

Many third party library exist to allow general purpose programmings
to access Redis databases through its TCP protocol. The most popular
such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.

The following program is how the previous example looks like in
Hedis:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 another-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34},\text{\tt \char34 c\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The function \ensuremath{\Varid{sadd}} takes a key and a list of values as
 arguments, and returns an \ensuremath{\Conid{Integer}} on success, or
 returns a \ensuremath{\Conid{Reply}}, a low-level representation of replies
 from the Redis server, in case of failures. All wrapped in
 \ensuremath{\Conid{Redis}}\footnotemark, the context of command execution.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{sadd}{}\<[7]%
\>[7]{}\mathbin{::}\Conid{ByteString}{}\<[27]%
\>[27]{}\mbox{\onelinecomment  key}{}\<[E]%
\\
\>[7]{}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]{}\<[27]%
\>[27]{}\mbox{\onelinecomment  values}{}\<[E]%
\\
\>[7]{}\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Note that keys and values, being nothing but binary strings in Redis,
are represented using Haskell \ensuremath{\Conid{ByteString}}.
Values of other types must be encoded as \ensuremath{\Conid{ByteString}}s before being written to the database,
and decoded after being read back.

\footnotetext{Hedis provides another kind of context, \ensuremath{\Conid{RedisTx}}, for \emph{transactions}, united with \ensuremath{\Conid{Redis}} under the class of \ensuremath{\Conid{RedisCtx}}. For brevity, we demonstrate only \ensuremath{\Conid{Redis}} in this paper. }

>>>>

\section{Motivation}
\label{sec:motivation}

All binary strings are equal, but some binary strings are more equal than others.

Although everything in Redis is essentially a binary string, these strings are treated differently. Redis supports many different
kind of data structures, such as strings, hashes, lists, etc. While
they are all encoded as binary strings before being written to the databse, most commands, much like how the C language treats a piece of data, only works with data of certain types.

\paragraph{Problem 1} The command \text{SET}, by
its definition, associates a key to a string. In the following
example, the key \text{some-string} is associated
to string \text{foo}. Subsequent calls to
\text{SADD} causes runtime errors, since the value
of \text{some-string} is not a set, but a string.

\begin{tabbing}\tt
~redis\char62{}~SET~some\char45{}string~foo\\
\tt ~OK\\
\tt ~redis\char62{}~SADD~some\char45{}string~bar\\
\tt ~\char40{}error\char41{}~WRONGTYPE~Operation~against~a~key\\
\tt ~~holding~the~wrong~kind~of~value
\end{tabbing}

\paragraph{Problem 2} Even worse, not all strings are equal!
The call \text{INCR some-string} parses the string
associated with key \text{some-string}
to an integer, increments it by one, and store it back as a string.
If the string can not be parse as an integer, a runtime error
is raised.

\begin{tabbing}\tt
~redis\char62{}~SET~some\char45{}string~foo\\
\tt ~OK\\
\tt ~redis\char62{}~INCR~some\char45{}string\\
\tt ~\char40{}error\char41{}~ERR~value~is~not~an~integer~or~out\\
\tt ~~of~range
\end{tabbing}

\paragraph{In Hedis} Hedis, being only a simple wrapper on top
of the TCP protocol of Redis, inherits all the problems mentioned
above. The following program yields the same error as that in
the Redis client.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{set}\;\text{\tt \char34 some-string\char34}\;\text{\tt \char34 foo\char34}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-string\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{Left}\;(\Conid{Error}\;\text{\tt \char34 WRONGTYPE~Operation~against~a\;~key~holding~the~wrong~kind~of~value\char34}{}\<[E]%
\\
\>[B]{}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\paragraph{The Cause} Every key is associated with a value, and every value has
 it's own type. But most commands in Redis only work with a certain type of
 value. When a command is used on a wrong type of key, a runtime error occurs.
 The problems illustrated above arise from the absence of type checking, with
 respects to \textbf{the type of a value that associates with a key}.
 These problems could have been avoided, if we could know the type every key
 associates with in advance, and prevent programs with invalid commands from
 executing.

\subsection{Hedis as an embedded DSL}

Haskell makes it easy to build and use Domain Specific Languages (DSLs),
 and Hedis can be regarded as one of them. What makes Hedis peculiar is that,
 it has \emph{variable bindings} (between keys and values), but with very
 little or no semantic checking, neither dynamically nor statically.

We began with making Hedis a dynamically typechecked embedded DSL, and implemented a
 runtime type checker that keeps track of types of all the variable. But then we
 found that things can be a lot easier, by leveraging the host language's type
 checker. We encode variable bindings with \emph{type-level lists} and
 \emph{strings}, and control the effects on the bindings with
 \emph{indexed monad}. In contrast to the former approach, we \textbf{embedded}
 our type checker into Haskell's type system, without having to build a
 \textbf{standalone} one on the term level.

\subsection{Contributions}

To summarize our contributions:

\begin{itemize}
\item We make Hedis statically type-checked, without runtime overhead.
\item We demonstrates how to model variable bindings of an embedded DSL with
 language extensions like type-level literals and data kinds.
\item We provide (yet another) an example of encoding effects and constraints of
 an action in types, with indexed monad\cite{indexedmonad} and other language
 extensions such as closed type-families\cite{closedtypefamilies} and
 constraints kinds\cite{constraintskinds}.
\item Popcorn, a package we built for programmers. This package helps programmers
 to write more reliable Redis programs, and also makes Redis polymorphic by
 automatically converting back and forth from values of arbitrary types and
 boring ByteStrings.
\end{itemize}
